---
title: "bimler_code_data_wrangle"
output: html_document
date: "2024-08-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# loading in data
```{r}
library(tidyverse)

# wir_focals_dat <- read.csv("~/Documents/REU/parks data/focals/wir fitness(WIR diverse focals).csv")
# wir_neighborhoods_data <- read.csv("~/Documents/REU/parks data/neighborhoods/wir fitness(WIR neighborhoods).csv")

# br_focals_dat <- read.csv("~/Documents/REU/parks data/focals/br fitness(BR diverse focals).csv")
# br_neighborhoods_data <- read.csv("~/Documents/REU/parks data/neighborhoods/br fitness(BR neighborhoods).csv")

sem_focals_dat <- read.csv("/sem fitness(SEM diverse focals.csv")
sem_neighborhoods_data <- read.csv("~/Documents/REU/parks data/neighborhoods/sem fitness(sem neighborhoods).csv")


# rf_focals_dat <- read.csv("~/Documents/REU/parks data/focals/riverfront fitness(Riverfront Diverse focals).csv")
# rf_neighborhoods_data <- read.csv("~/Documents/REU/parks data/neighborhoods/riverfront fitness(RIVERFRONT neighborhoods.csv")

```

# fixing column names for neighborhood data
```{r}
fix_col_names <- function(data){
  #setting column names to values of 1st row
  colnames(data) <- data [1, ]
  
  # removing 1st row
  data <- data[-1,]
  
  return (data)
}

# wir_neighborhoods_data <- fix_col_names(wir_neighborhoods_data)
sem_neighborhoods_data <- fix_col_names(sem_neighborhoods_data)
# rf_neighborhoods_data <- fix_col_names(rf_neighborhoods_data)
# br_neighborhoods_data <- fix_col_names(br_neighborhoods_data)

# filtering the neighborhoods so that we keep only the ones that have focal collects 
  # to be merged with the focal data

#unique(sem_neighborhoods_data$`focal collect?`)
sem_neighbors_filtered <- sem_neighborhoods_data[sem_neighborhoods_data$`focal collect?` == "Y", ]


```

# cleaning the fruit counts into one column
```{r}
# doing the fruit count
calculate_new_fruit_count <- function(row, data){
  # calculating the average side and tops in the function?
  
  average_top <- mean(data$TOPFR, na.rm = TRUE)
  average_side <- mean(data$SIDEFR, na.rm = TRUE)
  
  if(!is.na(data[row, 'TOPFR']) & !is.na(data[row, 'SIDEFR'])){
    return((data[row, 'TOPFR']) + ((data[row, "TOTALFR"] - 1) * data[row, 'SIDEFR']))
  }
  else{
    return(average_top + ((data[row, 'TOTALFR'] - 1) * average_side))
  }
}

# only want to apply the function to NAV and COLLOM 
nav_and_collom <- c("NAVSQU", "COLLOM")
# nav_and_collom_alone <-  wir_focals_dat[wir_focals_dat$SPECIES %in% nav_and_collom, ]
# 
# # applying function to the filtered nav/collom dataset
# nav_and_collom_alone$average_fruits <- sapply(1:nrow(nav_and_collom_alone), function(row) calculate_new_fruit_count(row, nav_and_collom_alone))
# 
# # Merge the new column back to the original dataframe
# wir_focals_dat <- merge(wir_focals_dat, nav_and_collom_alone[, c("SPECIES", "average_fruits")], by = "SPECIES", all.x = TRUE)
# 
# wir_focals_dat <- wir_focals_dat %>%
#   mutate(fruits = coalesce(average_fruits, EXACTFR, TOTALFR))
```

# SEM fixing fruit counts 
```{r}
###
# modifying TOTAL.FR to TOTALFR for fxn to work
  # new name = old name
sem_focals_dat <- sem_focals_dat %>% rename(TOTALFR = TOTAL.FR)


nav_and_collom_alone_sem <-  sem_focals_dat[sem_focals_dat$SPECIES %in% nav_and_collom, ]

# applying function to the filtered nav/collom dataset
nav_and_collom_alone_sem$average_fruits <- sapply(1:nrow(nav_and_collom_alone_sem), function(row) calculate_new_fruit_count(row, nav_and_collom_alone_sem))

# coalesce fruits into fruits
nav_and_collom_alone_sem <- nav_and_collom_alone_sem %>%
  mutate(fruits = coalesce(average_fruits, EXACTFR, TOTALFR))

# drop the avg fruits column
  #DF[ , !(names(DF) %in% drops)]

nav_and_collom_alone_sem <- select(nav_and_collom_alone_sem, -("average_fruits"))

# Merge the new column back to the original dataframe
# sem_focals_dat <- merge(sem_focals_dat, nav_and_collom_alone_sem[, c("SPECIES", "average_fruits")], by = "SPECIES", all.x = TRUE)
# trying a new way
# removing all nav and collom instances
sem_focals_dat_no_nav_or_collom <- sem_focals_dat[!(sem_focals_dat$SPECIES %in% c("NAVSQU", "COLLOM")), ]
# coalesce fruits into one fruits
sem_focals_dat_no_nav_or_collom <- sem_focals_dat_no_nav_or_collom %>%
  mutate(fruits = coalesce(EXACTFR, TOTALFR))

# now merging the nav_and_collom_alone_sem with sem_focals_dat_no_nav_or_collom
sem_updated_focals <- rbind(sem_focals_dat_no_nav_or_collom, nav_and_collom_alone_sem)

# print(colnames(sem_focals_dat_no_nav_or_collom))
# print(colnames(nav_and_collom_alone_sem))


```


# RF changing fruit counts (skipping for time being to focus on getting 1 complete datafrme for SEM)
```{r}
###############
### RF -- need to also changing FR1 to TOTAL FR
#############



# nav_and_collom_alone <-  rf_focals_dat[rf_focals_dat$SPECIES %in% nav_and_collom, ]
# 
# # applying function to the filtered nav/collom dataset
# nav_and_collom_alone$average_fruits <- sapply(1:nrow(nav_and_collom_alone), function(row) calculate_new_fruit_count(row, nav_and_collom_alone))
# 
# # Merge the new column back to the original dataframe
# rf_focals_dat <- merge(rf_focals_dat, nav_and_collom_alone[, c("SPECIES", "average_fruits")], by = "SPECIES", all.x = TRUE)
# 
# # combining fruits columns into one
# rf_focals_dat <- rf_focals_dat %>%
#   mutate(fruits = coalesce(average_fruits, EXACTFR, FR1))
```
#br changing fruit counts (skipping for time being to focus on getting 1 complete datafrme for SEM)
```{r}
# # creating a new column to use the fruit counts instead of the inflorescence count
# br_focals_dat$average_fruits <- rowMeans(br_focals_dat[, c('TOPFR', 'SIDEFR')], na.rm = TRUE)
# # combining fruits columns into one
# br_focals_dat <- br_focals_dat %>%
#   mutate(fruits = coalesce(average_fruits, EXACTFR, TOTAL.FR))

```


# merging each respective focal df with neighborhood df on plot/subplot
```{r}

# neighborhood_focal_wir <- merge(wir_neighborhoods_data, wir_focals_dat, by = c("PLOT.1", "SUBPLOT"))


# want to merge on all common columns to reduce chances of repeats?

common_cols <- intersect(colnames(sem_focals_dat), colnames(sem_neighborhoods_data))

print(colnames(sem_focals_dat))
print(colnames(sem_neighborhoods_data))
print(common_cols)

neighborhood_focal_sem <- merge(sem_neighbors_filtered, sem_updated_focals, by = c(common_cols))



# neighborhood_focal_rf <- merge(rf_neighborhoods_data, rf_focals_dat, by = c("PLOT.1", "SUBPLOT"))
# neighborhood_focal_br <- merge(br_neighborhoods_data, br_focals_dat, by = c("PLOT.1", "SUBPLOT"))

```


```{r}
# for some reason I'm finding a bunch of duplicate rows? NVM
# #new_df <- df[!duplicated(df), ]
# no_dup_neighborhood_focal_sem <- neighborhood_focal_sem[!duplicated(neighborhood_focal_sem),]
#  

```


# dropping/selecting columns we need
## focal species, seeds/fruits, all of the neighbor counts columns (for SEM)
```{r}
# KEEPING A PLOT COLUMN

neighborhood_focal_sem <- neighborhood_focal_sem %>% select(-PLOT.1, -SUBPLOT, -`focal collect?`, -PARK, -DATE, -REP, -HEIGHT, -STEMS, -TOTALFR, -BAGGED., -EXACTFR, -BIOMASS, -TOPFR, -SIDEFR, -NOTES )

# renaming PLOT.y to PLOT to be able to merge with the alones
#neighborhood_focal_sem <- neighborhood_focal_sem %>% rename(PLOT )


# need to alphabetize 
neighborhood_focal_sem <- neighborhood_focal_sem %>% arrange(SPECIES)
# changing SPECIES to focals
neighborhood_focal_sem <- neighborhood_focal_sem %>% rename(focal = SPECIES)
# (not necc correct) BUT changing fruits to seeds
neighborhood_focal_sem <- neighborhood_focal_sem %>% rename(seeds = fruits)

# dropping any rows for now that dont have a final_fruit_count
neighborhood_focal_sem <- neighborhood_focal_sem %>% filter(!is.na(seeds))

# trying to alphabetize the weed columns while leaving out seeds and focal

# Specify the columns to leave out or move to the front
columns_to_exclude <- c("focal", "seeds", "PLOT")
# Alphabetize the remaining columns
remaining_columns <- setdiff(names(neighborhood_focal_sem), columns_to_exclude)
sorted_columns <- remaining_columns[order(remaining_columns)]
# Combine the specified columns with the sorted columns
new_order <- c(columns_to_exclude, sorted_columns)
# Reorder the data frame
neighborhood_focal_sem <- neighborhood_focal_sem %>% select(all_of(new_order))

# filling any blanks with 0s
neighborhood_focal_sem[neighborhood_focal_sem == ''] <- 0

# in the case we dont need the alones

#write.csv(neighborhood_focal_sem, "only_diverse_sem.csv")

```

# adding in the alones 
## zeroing out in weeds, seed counts, species

# also keep the PLOT column for alones, then for the rest of the entries mark them as diverse?
```{r}
all_alones <- read.csv("~/Documents/REU/parks data/fitness alone focals(ALL alone focals).csv")
all_alones<- all_alones %>% rename(TOTALFR = fr1)


# doing the fruit count
calculate_new_fruit_count <- function(row, data){
  # calculating the average side and tops in the function?
  
  average_top <- mean(data$TOPFR, na.rm = TRUE)
  average_side <- mean(data$SIDEFR, na.rm = TRUE)
  
  if(!is.na(data[row, 'TOPFR']) & !is.na(data[row, 'SIDEFR'])){
    return(data[row, 'TOPFR'] + ((data[row, "TOTALFR"] - 1) *data[row, 'SIDEFR']))
  }
  else{
    return(average_top + ((data[row, 'TOTALFR'] - 1) * average_side))
  }
}

# only want to apply the function to NAV and COLLOM 
nav_and_collom <- c("NAVSQU", "COLLOM")
nav_and_collom_alone <-  all_alones[all_alones$SPECIES %in% nav_and_collom, ]

# applying function to the filtered nav/collom dataset
nav_and_collom_alone$NEW_FRUIT_COUNT <- sapply(1:nrow(nav_and_collom_alone), function(row) calculate_new_fruit_count(row, nav_and_collom_alone))

# Merge the new column back to the original dataframe
all_alones <- merge(all_alones, nav_and_collom_alone[, c("SPECIES", "NEW_FRUIT_COUNT")], by = "SPECIES", all.x = TRUE)


# Add missing columns to all_alones with zeros (for SEM)
missing_cols <- setdiff(names(neighborhood_focal_sem), names(all_alones))
all_alones[missing_cols] <- 0

# coalesce for new_fruit_count, then EXACTFR, then TOTALFR
all_alones <- all_alones %>%
  mutate(FINAL_FRUIT_COUNT = coalesce(NEW_FRUIT_COUNT, EXACTFR, TOTALFR))

# getting rid of columns we don't want 
  # want to keep all neighbor species, fr count, species, and PLOT

all_alones <- all_alones %>% select(-date, -height, -stemcount, -TOTALFR, -fr2, -fr3, -fr4, -fr5, - EXACTFR, -BIOMASS, -TOPFR, -SIDEFR, -notes, -NEW_FRUIT_COUNT, -focal, -seeds)

# now filtering for SEM 
all_alones_sem <- all_alones %>% filter (PARK=="SEM")

# dropping park and rep
all_alones_sem <- all_alones_sem %>% select (-REP, - PARK)

# dropping any rows for now that dont have a final_fruit_count
all_alones_sem <- all_alones_sem %>% filter(!is.na(FINAL_FRUIT_COUNT))


# renaming final_fruit count seeds
# need to alphabetize 
all_alones_sem <- all_alones_sem %>% arrange(SPECIES)
# changing SPECIES to focals
all_alones_sem <- all_alones_sem %>% rename(focal = SPECIES)
# (not necc correct) BUT changing fruits to seeds
all_alones_sem <- all_alones_sem %>% rename(seeds = FINAL_FRUIT_COUNT)

# trying to alphabetize the weed columns while leaving out seeds and focal

# Specify the columns to leave out or move to the front
columns_to_exclude <- c("focal", "seeds", "PLOT")
# Alphabetize the remaining columns
remaining_columns <- setdiff(names(all_alones_sem), columns_to_exclude)
sorted_columns <- remaining_columns[order(remaining_columns)]
# Combine the specified columns with the sorted columns
new_order <- c(columns_to_exclude, sorted_columns)
# Reorder the data frame
all_alones_sem <- all_alones_sem %>% select(all_of(new_order))

```
# SEM finalized dataset
```{r}
sem_neighbor_focal_all <- rbind(neighborhood_focal_sem, all_alones_sem)

# before saving it as a csv,for stan data we have to replace all nas with 0s!
  

# saving it as a csv!
#write.csv(sem_neighbor_focal_all, "sem_neighbor_focal_alone_diverse_merge.csv")

```
